Maximum Entropy Spectral Analysis: a case study

### 2021-03-25 paper/mesa.tex
Added and fixed refs

An ubiquitous method for such computation is due to Welch \cite{Welch} and it based on eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
Basically, the PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many windowed batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent \cite{Lomb,Scargle} to taking the Fourier Transform of the sample autocorrelation $\rho$, written as
\begin{equation}
    \rho_{Windowed} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different way, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

An appealing alternative approach, based on the Maximum Entropy principle \cite{Jaynes}, has been put forward by Burg \cite{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the arbitrary parameters, as it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provide a natural and simple machinery to forecast a time series, thus predicting past observations based on previous ones.

@@

}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalised by the discrete Fourier transform base\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i}\tilde{x}_i^2S^{-1}_{i} \right)\,,
\end{align}
where the matrix $S = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\cite{woodward}
%P. M. Woodward, D. W. Fry, and W. Higinbotham, Probability and information theory, with applications to radar; 2nd ed. (Pergamon Press, Oxford, 1964), URL http://cds.cern.ch/record/2031792.
and of gravitational waves data analysis, e.g. \cite{finn:1992}.
% L. S. Finn, Phys. Rev. D46, 5236 (1992), gr-qc/9209010
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the auto-covariance function.

@@

It turns out that the maximums entropy principle provides a representation of the time series as an $AR(p)$ process. The Burg's algorithm computes the autoregressive coefficients that are suitable to the available data.
Thanks to Wold's theorem \cite{something}, every stationary time series can be represented as an autoregressive process: this ensures that maximum entropy estimation is faithful and general.

@@

\subsection{Choice of the loss function: Gaussian PSD} \label{sec:arp_validation}
We test the performance of the three loss functions on a random time series generated with a known power spectral density.
In this first experiment, we take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$, see Fig.~\ref{fig:autocorr}.
The time series are generated by sampling a frequency vector from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$, following the procedure outlined in~\cite{}, and performing an inverse Fast Fourier Transform on the frequency series. 
For this investigation, we generate a dataset of 1000 time series of 3000 points each, see Fig.~\ref{fig:noise} for an example of a time series in frequency and time domain.

@@

\begin{figure}
	\centering
	\includegraphics[width = \linewidth]{Images/optimisers_comparison/normal/error_length_contour.pdf}
	\caption{Relative error $r$ (as in eq.~\eqref{eq:freq_error}) vs the length $m$ of autoregressive process, for each of the $1000$ independent realizations of the synthetic time series with gaussian PSD given in Fig.~\ref{fig:autocorr}. Different colours refers to different choices for the loss function: in red $\mathcal{L}_{\rm FPE}(m)$, in green $\mathcal{L}_{\rm OBD}(m)$ and in black $\mathcal{L}_{\rm CAT}(m)$. The top and right histograms show the marginal distributions.}
	\label{fig:optcomparison}
\end{figure}

@@

We perform two sets of experiments.
In the first one (see fig.~), we simulate data from an analytical PSD and we make the comparison with them: this is to ensure the we have a baseline PSD to compare the data with.
In a second experiment (see fig.~), we employ the data released to the public byt the LIGO/Virgo collaboration.
In each experiment, we vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we employ FPE optimizer. For the Welch algorithm, we employ a Tukey window, an overlap fraction of $0.5$ for the segments and a length of segments $L = 512, 1024, 2048, 8192, 32768$ datapoints.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.

@@

\begin{figure}
	\label{fig:mem_welch_realdata}
	\caption{Make here a comparison between Welch - True data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_overall_fake_False.pdf}
\end{figure}
\begin{figure}
	\label{fig:mem_welch_fakedata}
	\caption{Make here a comparison between Welch - Fake data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_overall_fake_True.pdf}
\end{figure}

@@

\section{Applications} \label{sec:applications}
\subsection{Temperature Time Series} \label{sec:temp_timeseries}
We apply our code to the study of the temperature time series. This will allow to test the code performance on a different time series, of great importance in physics. We will also assess the quality of the forecasting predictions: this task has a great impact also on practical applications.
We downloaded the historical reanalysis data from the Copernicus project (or whathever it is) \cite{}.
Data refers to the temperature measure at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperature is measured once per hour for almost 31 years from 31st December 1989 to 30th November 2020.

In fig.~\ref{fig:temp_spectrum}, we compute the spectrum over all the data. We employ both FPE and CAT loss functions.
First of all, we note that at every frequency the estimation given by FPE is much less noisy the one from CAT: this issue was already studied in the previous sections. 
We clearly see that there is a peak on the PSD at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to a day. This is somehow trivial and highly expected.
Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they are required to explain the high variability of the daily temperature cycle throughout the year.

@@

\subsection{Seismic waveforms} \label{sec:seismic_timeseries}
\blindtext
\begin{figure}
	\caption{Do a caption}
	\label{fig:seismic_spectrum}
	\includegraphics{Images/seismic_plots/seismic_spectrum.pdf}
\end{figure}

@@

	\begin{acknowledgments}
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center (https://www.gw-openscience.org/ ), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain.
	\end{acknowledgments}

>>>>

An ubiquitous method for such computation is due to Welch \cite{Welch1967} and it based on eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
Basically, the PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many windowed batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent \cite{Lomb,Scargle} to taking the Fourier Transform of the sample autocorrelation $\rho$, written as
\begin{equation}
    \rho_{Windowed} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different way, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

An appealing alternative approach, based on the Maximum Entropy principle \cite{JaynesArticle,jaynes2003ptl}, has been put forward by Burg \cite{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the arbitrary parameters, as it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provide a natural and simple machinery to forecast a time series, thus predicting past observations based on previous ones.

@@

}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalised by the discrete Fourier transform base\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i}\tilde{x}_i^2S^{-1}_{i} \right)\,,
\end{align}
where the matrix $S = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\cite{prob_information_theory}
%P. M. Woodward, D. W. Fry, and W. Higinbotham, Probability and information theory, with applications to radar; 2nd ed. (Pergamon Press, Oxford, 1964), URL http://cds.cern.ch/record/2031792.
and of gravitational waves data analysis, e.g. \cite{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the auto-covariance function.

@@

It turns out that the maximums entropy principle provides a representation of the time series as an $AR(p)$ process. The Burg's algorithm computes the autoregressive coefficients that are suitable to the available data.
Thanks to Wold's theorem \cite{Wold_theorem}, every stationary time series can be represented as an autoregressive process: this ensures that maximum entropy estimation is faithful and general.

@@

\subsection{Choice of the loss function: Gaussian PSD} \label{sec:arp_validation}
We test the performance of the three loss functions on a random time series generated with a known power spectral density.
In this first experiment, we take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$.
The time series are generated by sampling a frequency vector from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$, following the procedure outlined in sec.~\ref{sec:forecasting}, and performing an inverse Fast Fourier Transform on the frequency series. 
For this investigation, we generate a dataset of 1000 time series of 3000 points each.

@@

\begin{figure}
	\centering
	\includegraphics[width = \linewidth]{Images/optimisers_comparison/normal/error_length_contour.pdf}
	\caption{Relative error $r$ (as in eq.~\eqref{eq:freq_error}) vs the length $m$ of autoregressive process, for each of the $1000$ independent realizations of the synthetic time series with a gaussian PSD. Different colours refers to different choices for the loss function: in red $\mathcal{L}_{\rm FPE}(m)$, in green $\mathcal{L}_{\rm OBD}(m)$ and in black $\mathcal{L}_{\rm CAT}(m)$. The top and right histograms show the marginal distributions.}
	\label{fig:optcomparison}
\end{figure}

@@

We perform two sets of experiments.
In the first one (see fig.~\ref{fig:mem_welch_fakedata}), we simulate data from an analytical PSD and we make the comparison with them: this is to ensure the we have a baseline PSD to compare the data with.
In a second experiment (see fig.~\ref{fig:mem_welch_realdata}), we employ the data released to the public by the LIGO/Virgo collaboration \cite{Abbott_2021}.
In each experiment, we vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we employ FPE optimizer. For the Welch algorithm, we employ a Tukey window, an overlap fraction of $0.5$ for the segments and a length of segments $L = 512, 1024, 2048, 8192, 32768$ datapoints.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.

@@

\begin{figure}
	\label{fig:mem_welch_fakedata}
	\caption{Make here a comparison between Welch - Fake data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_overall_fake_True.pdf}
\end{figure}
\begin{figure}
	\label{fig:mem_welch_realdata}
	\caption{Make here a comparison between Welch - True data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_overall_fake_False.pdf}
\end{figure}

@@

\section{Applications} \label{sec:applications}
\subsection{Temperature Time Series} \label{sec:temp_timeseries}
We apply our code to the study of the temperature time series. This will allow to test the code performance on a different time series, of great importance in physics. We will also assess the quality of the forecasting predictions: this task has a great impact also on practical applications.
We used the historical reanalysis data from the ``ERA5-Land hourly data from 1981 to present" dataset, downloaded from the Climate Data Store \cite{ERA5_dataset}

Data refers to the temperature measure at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperature is measured once per hour for almost 31 years from 31st December 1989 to 30th November 2020.

In fig.~\ref{fig:temp_spectrum}, we compute the spectrum over all the data. We employ both FPE and CAT loss functions.
First of all, we note that at every frequency the estimation given by FPE is much less noisy the one from CAT: this issue was already studied in the previous sections. 
We clearly see that there is a peak on the PSD at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to a day. This is somehow trivial and highly expected.
Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they corresponds to signals that preserve the 1 day period while, at the same time, capturing the complex variability of the daily temperature cycle throughout the year.

@@

\subsection{Seismic waveforms} \label{sec:seismic_timeseries}
\blindtext

\cite{seismic_data}

\blindtext

\begin{figure}
	\caption{Do a caption}
	\label{fig:seismic_spectrum}
	\includegraphics{Images/seismic_plots/seismic_spectrum.pdf}
\end{figure}

@@

	\begin{acknowledgments}
\sschimidt{Is it fine to ackonowledge people helping with temperature data and seismic data?}
%S.S. wants to thank Maurizio Maugeri and Costanza Rossi for useful discussions.

This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center (\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain.
	\end{acknowledgments}
